Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-17 Manuscript/manuscript.Rmd
editted intro

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. In doing so, many methods ignore key characteristics of domain data that can impair and bias results. Most notably is the introduction of heavily imbalanced classes. Due to the sparcity of TADs throughout the genome, most defined regions will not contain a TAD boundary. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data. The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance.

@@

A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. Here, the authors ignored the effect of imbalanced classes as well as the use of potentially biased performance metrics.

In a similar study, Hong et al. used domain data from an hESC cell line to build a linear model and predict which genomic elements were associated with TAD boundaries [@hong2017computational]. They defined boundary segments as genomic regions before and after 150 kilobases of the boundary. To solve the imbalance problem, they randomly sampled from regions outside of the 150kb flank.  For their model, the feature space was determined by the number of counts of ChIP-seq peaks defining functional genomic elements that overlapped with the two segments. Individual models were then performed until the best model which included the concensus CTCF signal feature, followed by 10 additional elements was obtained. Here, however, it is likely that 1 iteration of random under-sampling will yield biased results, depending on how representative the under-sampled data is. Likewise, it is unclear how the performances of other balancing techniques compare (i.e. over-sampling or a combination of both under and over-sampling).

@@

The coordinates of the corresponding 10 kb resolution contact maps resulted in a total of X genomic bins [??? To be modified for the whole genome]. There were only Y bins that overlapped with a flanked TAD boundary representing the minority class, for an imbalance ratio of (X-Y):Y. After applying each resampling technique to the training set, we obtained perfectly balanced classes for all except the ROSE technique (Figure 4).

>>>>

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not [@mourad2016computational, @hong2017computational]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data that can impair and bias results.

Most notably is the introduction of heavily imbalanced classes. Due to the sparcity of TADs throughout the genome, most defined regions will not contain a TAD boundary. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to large imbalance ratios (defined as the ratio of majority to minority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple, @chen2013influence]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include the area under the receiver operating curve (AUROC). The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data in the area under the precision-recall curve (AUPRC). It has been shown that the use of AUPRC with imbalanced datasets yields more accurate and intuitive interpretation of model performance when compared to AUROC [@saito2015precision]. Likewise, metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data. In each case, the weight

do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance. Better measures

@@

?????????????????????????????????????????????????????????????????????????[include if we decide to directly compare these methods; otherwise referenced above]
A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. Here, the authors ignored the effect of imbalanced classes as well as the use of potentially biased performance metrics.

In a similar study, Hong et al. used domain data from an hESC cell line to build a linear model and predict which genomic elements were associated with TAD boundaries [@hong2017computational]. They defined boundary segments as genomic regions before and after 150 kilobases of the boundary. To solve the imbalance problem, they randomly sampled from regions outside of the 150kb flank.  For their model, the feature space was determined by the number of counts of ChIP-seq peaks defining functional genomic elements that overlapped with the two segments. Individual models were then performed until the best model which included the concensus CTCF signal feature, followed by 10 additional elements was obtained. Here, however, it is likely that 1 iteration of random under-sampling will yield biased results, depending on how representative the under-sampled data is. Likewise, it is unclear how the performances of other balancing techniques compare (i.e. over-sampling or a combination of both under and over-sampling).
?????????????????????????????????????????????????????????????????????????

@@

The coordinates of the corresponding 10 kb resolution contact maps resulted in a total of X genomic bins [??? To be modified for the whole genome]. There were only Y bins that overlapped with a flanked TAD boundary representing the minority class, for an imbalance ratio of (X-Y):Y [??? To be modified for the whole genome]. After applying each resampling technique to the training set, we obtained perfectly balanced classes for all except the ROSE technique (Figure 4).
